---
title: "FAQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Authentication

#### Q-Auth-1

I am getting an error when using one of `travis::use_travis_deploy()`, `circle::use_circle_deploy()` or `tic::use_ghactions_deploy()`.

**Answer**

In most cases this is related to API authentication issues.
Ensure that the following points are met:

1. For Circle CI and Travis CI, install the respective GitHub App from the [GitHub Marketplace](https://github.com/marketplace).
2. Ensure that you have set the respective API keys for the problematic provider in your `.Renviron` file.
  Consult the help pages of the respective `use_*_deploy()` function for more help.
  - GitHub Actions: A `GITHUB_PAT` with "public_repo" scopes.
  - Travis CI: Env var `R_TRAVIS_ORG` or `R_TRAVIS_COM`, depending on the endpoint you use.
  - Circle CI: Env var `R_CIRCLE`.

## GitHub Actions

#### Q-GHA-1

How is {tic} different from what [r-lib/actions](https://github.com/r-lib/actions) does?

**Answer**

{tic} uses [r-lib/actions](https://github.com/r-lib/actions) as the base to install R in the first place.
However in detail, {tic} does the following things differently which aim to enhance the CI experience:

- Caching: {tic} caches the whole R library rather than only the direct dependendencies of a package. 
  This has the advantage that packages required for side actions ({pkgdown} deployment, README updates) will also be cached.
  
- `ccache`: {tic} comes with a compiler cache for source installations of packages by default, speeding up repeated source installation highly.
  The compiler cache directory (`.ccache`) will also be cached (once a week).
  Example use case: If you installed {Rcpp} from source as a dependency of your package and have it stored in your cache and {Rcpp} now updates two days later, the reinstallation will make use of the compiler cache and install {Rcpp} instantly rather than recompiling the C code again. 
  
- Number of CPUs: {tic} uses 4 CPUs by default instead of only 1 as [r-lib/actions](https://github.com/r-lib/actions) does.
  This speeds up package installation a lot.
  4 CPUs are max because all GitHub Actions runners have 2 hyperthreading cores available.
  
- Use of SSH deployment keys: Once set up via `tic::use_ghactions_deploy()`, this deployment approach makes it possible to push any file of your repository to any branch of your remote.
  Other deployment approaches often limit you to only push to `gh-pages` branch or similar.
  
## Travis CI

#### Q-Travis-1

What is the difference between travis-ci.org and travis-ci.com and which one should I use?

**Answer**

When dealing with Travis CI (or the [{travis}](https://github.com/ropenscilabs/travis) package) the first time, one might be confused what the difference is between https://travis-ci.org and https://travis-ci.com.
[This devops.stackexchange question](https://devops.stackexchange.com/questions/1201/whats-the-difference-between-travis-ci-org-and-travis-ci-com) gives some insights on the history of both services and why both still exist.

- Both https://travis-ci.org and https://travis-ci.com are the same whereas the future belongs to https://travis-ci.com.
- Due to both being still active, the [{travis}](https://github.com/ropenscilabs/travis) package supports both endpoints for API queries.
- If you start a new project, we recommend to enable the repo on https://travis-ci.com. 
  If you are using the {travis] package, the default endpoint is set to `".com"` so that all function call will try to interact with repositories on https://travis-ci.com.
  This can be changed via env var `R_TRAVIS`.
  
#### Q-Travis-2

I am seeing conflicts when deploying a pkgdown site during deployment - it looks like multiple runners are facing a git race condition.
I am getting errors like

```sh
HEAD is now at ada17f3 Deploy from Travis build 13 [ci skip]
git cherry-pick --no-commit 8f4e52da03a6ccdb66a140c3ca08be22d57d7288
error: could not apply 8f4e52d... Deploy from Travis build 13 [ci skip]
hint: after resolving the conflicts, mark the corrected paths
hint: with 'git add <paths>' or 'git rm <paths>'
```

**Answer**

Correct. 
This is because multiple runners building and deploying the pkgdown site nearly at the same time.
To prevent this from happening (doing this for one runner is enough) condition the pkgdown step to one runner by setting an env var.

In `tic.R`:

```r
if (ci_on_travis() && ci_has_env("PKGDOWN") {
  # creates pkgdown site and pushes to gh-pages branch
  do_pkgdown()
}
```

and then set this env var for one of the runners in `.travis.yml`:

```yml
matrix:
  include:
    - os: osx
      r: release
    - os: linux
      r: release
      env:
      - PKGDOWN=true
```
  
## Appveyor CI

#### Q-Appveyor-1

## Appveyor

I am facing the following build error:

```{r eval=FALSE}
Error in eval(expr, envir = list(`?` = function(...) stop()), enclos = envir) : 
  object 'name' not found
Calls: <Anonymous> ... .transformer -> glue -> glue_data -> <Anonymous> -> .transformer
Execution halted
Command exited with code 1
```

**Answer**

Do you have a private SSH key stored as a secure environment variable in the in `appveyor.yml`?
If yes, it could be that you encrypted while being logged in to a different account than the one the repo is running under in reality.
This means if you have access to multiple organizations and your private account on Appveyor, you need to encrypt the environment variable using the account in which repo lives.

## Other

#### Q-Other-1

Is it possible to update the CI YAML templates installed by {tic} with upstream changes?

**Answer**

Not yet, you have to run `use_tic()` again and save your custom changes beforehand and reapply those.
We are planning to add an update mechanism soon though!

---

#### Q-Other-2

Am I the only one using {tic}?

**Answer**

You can see who and how many people use {tic.R} on GitHub via this query: https://github.com/search?p=5&q=filename%3Atic.R&type=Code

---

#### Q-Other-3

Package {rgl} fails to install because of either

- "configure: error: X11 not found but required, configure aborted."
- "error: X11 not found; XQuartz (from www.xquartz.org) is required to run rgl."

**Answer**

The first one is usually caused by a missing installation of `XQuartz` on macOS.
Add `brew cask install xquartz` to the runner.

The second error requires to set the `DISPLAY` env var to mimic a non-headless state.
Add `export DISPLAY=:99` to the stage in which {rgl} should be installed.
