---
title: "Getting started with CI for R"
author: "Patrick Schratz, Kirill Müller"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CI for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Prerequisites

If you are unfamiliar with the term Continuous Integration (CI), we highly recommend to read the following resources:

- https://ropensci.github.io/dev_guide/ci.html
- https://juliasilge.com/blog/beginners-guide-to-travis/
- http://mahugh.com/2016/09/02/travis-ci-for-test-automation/

See the [Features](features.html) article for a comparison with other solutions for Continuous Integration for R.


# Initialization/Setup

The easiest way to use {tic} for CI services is to call `tic::use_tic()`.
This will initialize a setup wizard which will guide you through all possibilities of the offered CI providers by {tic}.

Several yes/no questions need to be answered. 
Based on the replies we'll select specific templates.

Besides the question which CI system you want to use, you'll be asked if you want to 

- Deploy from builds to Github (e.g. if you are building a {pkgdown} site)

- Test your package on multiple R versions

```r
tic::use_tic()

→ Welcome to → tic!
This wizard will guide you through the setup process for getting started with various CI
providers.

Introduction:

tic currently comes with support for three CI providers:
● Appveyor
● Circle CI
● Travis CI

There is no need to use all of them. You can choose which one(s) you want to use, whether
you want to deploy (i.e. push from builds) and if you want to test on multiple R
versions.

We recommend the following setup:
● Appveyor: Windows
● Travis CI: Linux
● Travis CI: macOS

Selection: 2
Choosing your setup.

We'll ask you a few yes/no questions to gather your preferences.

Which provider do you want to use for Linux builds? 

1: Travis CI
2: Circle CI
3: None
4: All

Do you want to build on macOS (= Travis CI)? 

1: Travis CI
2: None

Do you want to build on Windows (= Appveyor)? 

1: AppVeyor CI
2: None

Do you want to deploy (i.e. push from the CI build to your repo) on certain providers? 
If yes, which ones? 

1: Travis CI
2: Circle CI
3: None
4: All

Do you want to build on multiple R versions? (i.e. R-devel, R-release, R-oldrelease). 
If yes, on which platform(s)? 

1: None
2: Travis CI
3: Circle CI
4: AppVeyor CI
5: All
```

Last, we'll add a `tic.R` file to the project root.

After this, your project is ready for continuous integration.
The next push to GitHub will create a build on Travis CI, Circle CI, and/or AppVeyor, as configured.
(Currently you need to enable the project manually on AppVeyor.)
See the [Troubleshooting](advanced.html#troubleshooting) section in case anything doesn't work as expected.

The reason why we suggest Travis CI as the default builds on three points:

- Travis is the most widely used CI provider within the R community
- Has been around for some years
- Supports Linux & macOS.

If you are open to try out new things, Circle CI comes with some advantages that might simplify your CI experience.
However, all providers come with pros and cons and we cannot provide an exhaustive list comparing all providers here.

The deployment setup functionality comes from the {travis} and {circle} packages.
See the [CI Client Packages](ci-client-packages.html) article for more detailed information on how {tic} and the CI client packages work together.

# The role of the `tic.R` file

After having called `tic::use_tic()` you will find

- `.travis.yml`
- `.circleci/config.yml`
- `appveyor.yml`
- `.github/workflows`

and a `tic.R` file in your repo, depending on the choices you made during `use_tic()`.
The latter will always be present because it will be the main CI config file for all providers from now on.
Usually you do not need to touch the YAML files anymore.
All build customization is done in `tic.R` and applies to all providers.
For more information about the build lifecycle in general, check the [Build lifecycle](build-lifecycle.html) article.

The basic `tic.R` template looks as follows:

```{r eval = FALSE}
do_package_checks()

if (ci_on_travis()) {
  do_pkgdown()
}
```

`tic.R` file has a declarative nature: It should consist of "stages", "steps" and "macro" functions (see below).
These functions will only have an effect when specified in `tic.R` and should **not** be used standalone as they will only run in a (simulated) CI run following a certain order.

To run plain R code within the build, encapsulate it within `add_code_step(<code>)` and add it to a certain build stage. 
See [the Build Lifecycle](build-lifecycle.html) article for detailed info about how to do this.

## Macros {#macros}

{tic} builds on the "macro" idea.
Macros are essentially wrappers of a sequence of steps for often used tasks on the CI system:

- Checking a package (R CMD check)
- Building and deploying a pkgdown site
- Build and deploy a bookdown project

They can be distinguished from other functions by their `do_` prefix.
If you have a good use case for a macro, let us know by opening an [issue](https://github.com/ropensci/tic/issues).

```{r eval = FALSE}
do_package_checks()
```

`do_package_checks()` adds essential steps to various stages of a CI run. 
Most importantly, it adds `step_rcmdcheck()` to the "script" stage. 
This step performs the check of an R package.
Afterwards, the code coverage is being checked using `covr::codecov()`.
See `?do_package_checks()` for more information.

```{r}
# step_install_deps() in the "install" stage, using the repos argument.
#
# step_rcmdcheck() in the "script" stage, using the warnings_are_errors, 
#  notes_are_errors, args, and build_args arguments.
#
# A call to covr::codecov() in the "after_success" stage (only if the codecov flag is set)
```

The other macro in the default template is `do_pkgdown()`.

```{r eval = FALSE}
if (ci_on_travis()) {
  do_pkgdown()
}
```

`do_pkgdown()` adds five steps to the build process:

```{reval = FALSE}
# step_install_deps() in the "install" stage, using the repos argument.
# 
# step_setup_ssh() in the "before_deploy" to setup the upcoming deployment (if deploy is set),
# 
# step_setup_push_deploy() in the "before_deploy" stage (if deploy is set),
# 
# step_build_pkgdown() in the "deploy" stage, forwarding all ... arguments.
# 
# step_do_push_deploy() in the "deploy" stage.
```

By default this currently happens only on Travis CI, because `ci_on_travis()` is used as a condition.
Why do we do this? Building the pkgdown site on multiple CI services has no added benefit and might even cause problems due to race conditions during deployment.

`ci_on_travis()` can be replaced by one of its sibling functions like `ci_on_ghactions()`.

# Example projects

{tic} can be used for various R projects.
Simply check out the setup for the type of project you are interested in to get a feeling of how {tic} is used in practice.

- [tic.package](https://github.com/ropenscilabs/tic.package): R packages with [{pkgdown}](https://pkgdown.r-lib.org/) documentation
- [tic.packagedocs](https://github.com/ropenscilabs/tic.packagedocs): R packages with [_packagedocs_](http://hafen.github.io/packagedocs/) documentation
- [tic.drat](https://github.com/ropenscilabs/tic.drat): CRAN-like package repositories with [_drat_](http://dirk.eddelbuettel.com/code/drat.html)
- [tic.blogdown](https://github.com/ropenscilabs/tic.blogdown): Blogs with [_blogdown_](https://bookdown.org/yihui/blogdown/)
- [tic.bookdown](https://github.com/ropenscilabs/tic.bookdown): Books with [_bookdown_](https://bookdown.org/)
- [tic.website](https://github.com/ropenscilabs/tic.website): Websites with [_rmarkdown_](https://rmarkdown.rstudio.com/)
- [tic.figshare](https://github.com/ropenscilabs/tic.figshare): Deploying artifacts to [figshare](https://figshare.com/) (work in progress).
- [tic.covrpage](https://github.com/ropenscilabs/tic.covrpage): Unit test summary report.

### Blogdown

As a show case, we explain a "blogdown" project in more detail.
[`blogdown`](https://bookdown.org/yihui/blogdown/) is an R package for publishing websites.
Under the hood, it uses the framework [Hugo](https://gohugo.io/) which gets installed by the respective `tic.R` [template](https://github.com/krlmlr/tic.blogdown/blob/975aedd43fec1dd55e8348eccfca2c7c5f663006/tic.R#L5) in the "install" section:

```{r eval = FALSE}
get_stage("install") %>%
  add_code_step(blogdown::install_hugo())
```

Next the website is built and deployed.
The `blogdown::build_site()` function for websites is the equivalent to `pkgdown::build_site()` for R packages.

```{r eval = FALSE}
get_stage("deploy") %>%
  add_code_step(blogdown::build_site()) %>%
  add_step(step_push_deploy())
```

Steps and stages differ between projects (e.g. between a "blogdown" website and a "package").
{tic} is smart enough to detect your project automatically when calling `tic::use_tic()` and will add the correct template.

**Note:** Currently, publishing to https://figshare.com/ doesn't work.
Also, publishing to https://zenodo.org/ is work in progress.

## {tic} projects from the community

The templates we provide with {tic} are minimal working examples.
In the following we list various R projects with non-standard/tailored {tic}-based workflows.
You may find inspiration for your own project there.
(Please contact us if you want to add yours to the list!)

- [mlr](https://github.com/mlr-org/mlr)
- [sf](https://github.com/r-spatial/sf)
- [dbi](https://github.com/r-dbi/DBI)
