---
title: "Getting started with CI for R"
author: "Patrick Schratz, Kirill Müller"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CI for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Prerequisites

If you are unfamiliar with the term Continuous Integration (CI), we highly recommend to read the following resources:

- https://ropensci.github.io/dev_guide/ci.html
- https://juliasilge.com/blog/beginners-guide-to-travis/
- http://mahugh.com/2016/09/02/travis-ci-for-test-automation/

See the ["Features" article](features.html) for a comparison with other solutions for Continuous Integration for R.


# Initialization

The easiest way to use *tic* for CI services is to call `tic::use_tic()`.
This will create templates for both "Travis CI" and "Appveyor" and initialize all the required authentication steps for deployment.
During the process, browser pages will open up, ensuring that all permissions are set correctly and all apps are authorized.
Also a Personal Access Token (PAT) will be created on GitHub and stored  as an encrypted variable on Travis.
This avoids hitting rate limitations when accessing the GitHub API, see [Locke Data's blog post about GitHUB PATs](https://itsalocke.com/blog/using-travis-make-sure-you-use-a-github-pat/) for more details.
The PAT creation cannot be fully automated as some input from the user on GitHub is required when creating the PAT.

First, authentication with Travis (a browser window opens) and GitHub is initiated (another browser window opens).
Also, the repo is activated on Travis.

```r
tic::use_tic()
┌───────────────────────┐
│                       │
│   Welcome to `tic`!   │
│                       │
└───────────────────────┘
ℹ This wizard will set all the required tokens and files
  on Travis CI and Github. Lets get started!
✔ Setting active project to '/home/pjs/git/test2'
ℹ Using active project '/home/pjs/git/test2'
┌──────────────────────────────────────────────────────┐
│                                                      │
│   Step #1: We check if a Github repository exists.   │
│                                                      │
└──────────────────────────────────────────────────────┘
❯  Authenticating to GitHub.
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
✔ Authentication successful.
❯  Authenticating to GitHub.
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
✔ Authentication successful.
✔ Github repo exists.
┌─────────────────────────────────────────────────────┐
│                                                     │
│   Step #2: We check if Travis is already enabled.   │
│                                                     │
└─────────────────────────────────────────────────────┘
❯  Authenticating to GitHub.
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
✔ Authentication successful.
❯  Authenticating to GitHub.
❯  Authenticating to GitHub.
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
✔ Authentication successful.
Finished activating repo pat-s/test2 on Travis CI.
```

Next, you will be asked if the current `.travis.yml` should be replaced with the template from *tic* (if one exists).
Otherwise, the template is added without asking.

```r
┌────────────────────────────────────────────────┐
│                                                │
│   Step #3: We create new files                 │
│   `.travis.yml`, `appveyor.yml` and `tic.R`.   │
│                                                │
└────────────────────────────────────────────────┘
Overwrite pre-existing file '.travis.yml'?
1: Absolutely not
2: Nope
3: Yeah
```

The same is done for Appveyor.

```
Selection: 3
✔ Writing '.travis.yml'
Overwrite pre-existing file 'appveyor.yml'?
1: Yeah
2: Nope
3: No way

Selection: 1
✔ Writing 'appveyor.yml'
```

And also for the `tic.R` file.
Additionally, it is added to `.Rbuildignore` to not interfer with the package building.

```
Overwrite tic.R? 

1: Yes
2: No

Selection: 1
Added tic.R from template.
✔ Adding '^tic\\.R$' to '.Rbuildignore'
```

Next, the private deploy key for Travis is added to the GitHub repo.
This is needed to give Travis permissions to deploy into the GitHub repo.

```r
┌───────────────────────────────────────────┐
│                                           │
│   Step #4: We create a SSH key pair       │
│   to allow Travis deployment to Github.   │
│                                           │
└───────────────────────────────────────────┘
Finished adding deploy keys on GitHub for repo <username/repo>.
✔ Successfully added public deploy key 'travis+tic' to GitHub for <username/repo>. 
❯  You should receive a confirmation e-mail from GitHub.
❯  Delete the key in the repositorys settings to revoke access for that key or when you no longer need it.
```

Besides the private deploy key, also the public deploy key needs to be added as an environment variable to Travis.
As it is a SSH key, its named `id_rsa` by default.

```r
❯  Authenticating to GitHub.
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
✔ Authentication successful.
Finished updating private environment variable id_rsa for <username/repo> on Travis CI.
✔ Successfully added private deploy key to <username/repo> as secure environment variable id_rsa to Travis CI.
```

Last, a PAT is created to avoid the download limit of GitHub packages on Travis.
You now see two environment variables in your setting on Travis CI: `GITHUB_PAT` and `id_rsa`.

```r
┌─────────────────────────────────────────────────────────┐
│                                                         │
│   Step #5: We create a Github PAT key on Travis CI      │
│   to avoid Github API rate limitations in the builds.   │
│                                                         │
└─────────────────────────────────────────────────────────┘
❯  Authenticating to GitHub.
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
✔ Authentication successful.
❯  Creating a personal access token (PAT).
❯  The suggested description 'travis+tic for pat-s/test2' has been copied to the clipboard.travis+tic for pat-s/test2
ℹ  If you use this token only to avoid GitHubs rate limit, you can leave all scopes unchecked.
ℹ  Then, copy the new token to the clipboard, it will be detected and applied automatically.
ℹ Please visit
  https://github.com/settings/tokens/new
  A browser window will be opened.
ℹ  If you use this token only to avoid GitHubs rate limit, you can leave all scopes unchecked.
```

Most functionality of the wizard comes from the _travis_ package.
See the ["tic & travis" article](tic-travis.html) for more detailed information on how *tic* and *travis* work together.

# Explanation of the basic template

After having called `tic::use_tic()` you will find a `.travis.yml`, `appveyor.yml` and a `tic.R` file in your repo.
Usually you do not need to touch `appveyor.yml` and `.travis.yml` anymore.
All build customizations are done in `tic.R` and apply to both services.
For more information about the whole build lifecycle, check the ["Build lifecycle" article](build-lifecycle.html).

The basic `tic.R` template looks as follows:

```{r eval = FALSE}
do_package_checks()

if (ci_on_travis()) {
  do_pkgdown()
}
```

Let's break down what happens here:

```{r eval = FALSE}
do_package_checks()
```

`do_package_checks()` Is a macro which adds essential steps to various stages of a CI run. 
Essentially, it adds `step_rcmdcheck()` to the "script" stage. 
This step performs the check of a R package.
Afterwards, the code coverage is beim checked using `covr::codecov()`.

```{r}
# step_install_deps() in the "install" stage, using the repos argument.
# 
# step_rcmdcheck() in the "script" stage, using the warnings_are_errors, notes_are_errors, args, and build_args arguments.
# 
# A call to covr::codecov() in the "after_success" stage (only if the codecov flag is set)
```

The other macro in the default template is `do_pkdown()`.

```{r eval = FALSE}
if (ci_on_travis()) {
  do_pkgdown()
}
```

`do_pkgdown()` adds five steps to the build process:

```{reval = FALSE}
# step_install_deps() in the "install" stage, using the repos argument.
# 
# step_setup_ssh() in the "before_deploy" to setup the upcoming deployment (if deploy is set),
# 
# step_setup_push_deploy() in the "before_deploy" stage (if deploy is set),
# 
# step_build_pkgdown() in the "deploy" stage, forwarding all ... arguments.
# 
# step_do_push_deploy() in the "deploy" stage.
```

In a nutshell: All dependencies are being installed, the deployment setup is initialized, the pkgdown site is build and deployed.
This happens only on Travis CI, because `ci_on_travis()` evaluates to `TRUE` only on that platform.

# Examples projects

*tic* can be used for various R projects:

- [tic.package](https://github.com/ropenscilabs/tic.package): R packages with [_pkgdown_](https://pkgdown.r-lib.org/) documentation
- [tic.packagedocs](https://github.com/ropenscilabs/tic.packagedocs): R packages with [_packagedocs_](http://hafen.github.io/packagedocs/) documentation
- [tic.drat](https://github.com/ropenscilabs/tic.drat): CRAN-like package repositories with [_drat_](http://dirk.eddelbuettel.com/code/drat.html)
- [tic.blogdown](https://github.com/ropenscilabs/tic.blogdown): Blogs with [_blogdown_](https://bookdown.org/yihui/blogdown/)
- [tic.bookdown](https://github.com/ropenscilabs/tic.bookdown): Books with [_bookdown_](https://bookdown.org/)
- [tic.website](https://github.com/ropenscilabs/tic.website): Websites with [_rmarkdown_](https://rmarkdown.rstudio.com/)
- [tic.figshare](https://github.com/ropenscilabs/tic.figshare): Deploying artifacts to [figshare](https://figshare.com/) (work in progress).
- [tic.covrpage](https://github.com/ropenscilabs/tic.covrpage): Unit test summary report.

As a show case, we explain a "blogdown" project in more detail.
[`blogdown`](https://bookdown.org/yihui/blogdown/) is an R package for publishing websites.
Under the hood, it uses the framework [HUGO](https://gohugo.io/) which gets installed by the respective `tic.R` [template](https://github.com/krlmlr/tic.blogdown/blob/975aedd43fec1dd55e8348eccfca2c7c5f663006/tic.R#L5) in the "install" section:

```{r eval = FALSE}
get_stage("install") %>%
  add_code_step(blogdown::install_hugo())
```

Then the website is built and deployed.
The `blogdown::build_site()` function for websites is the equivalent to `pkgdown::build_site()` for R packages.

```{r eval = FALSE}
get_stage("deploy") %>%
    add_code_step(blogdown::build_site()) %>%
    add_step(step_push_deploy())
```

Steps and stages differ between projects (e.g. between a "blogdown" website and a "package").
*tic* is smart enough to detect your project automatically when calling `tic::use_tic()` and will add the correct template.

**Note:** Currently, publishing to https://figshare.com/ doesn't work.
Also, publishing to https://zenodo.org/ is work in progress.

## *tic* based R projects from the community

The templates we provide with *tic* are minimal working examples.
In the following we list various R projects with non-standard/tailored *tic*-based workflows.
You may find inspiration for your own project there.
(Please contact us if you want to add yours to the list!)

- [mlr](https://github.com/mlr-org/mlr)
- [sf](https://github.com/r-spatial/sf)
- [dbi](https://github.com/r-dbi/DBI)


# Advanced

The advanced usage of *tic* is described in more detail in the ["Advanced Usage" article](advanced.html):

- [Options for _pkgdown_ deployment](advanced.html#pkgdown-deployment) 
- [Using Travis CI Meta-information](advanced.html#using-travis-ci-meta-information)
- [Troubleshooting: Running *tic* locally](advanced.html#troubleshooting-running-tic-locally)
- [Troubleshooting: Enter into the Travis build](advanced.html#troubleshooting-running-tic-locally)

The build lifecycle when using *tic* is explained in the ["Build lifecycle" article](build-lifecycle.html)e.
Implementation of custom steps is described in the ["Writing custom steps" article](custom-steps.html).
