---
title: "Getting started with CI for R"
author: "Patrick Schratz, Kirill Müller"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CI for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Prerequisites

If you are unfamiliar with the term Continuous Integration (CI), we highly recommend to read the following resources:

- https://ropensci.github.io/dev_guide/ci.html
- https://juliasilge.com/blog/beginners-guide-to-travis/
- http://mahugh.com/2016/09/02/travis-ci-for-test-automation/

The most important improvements over existing solutions are:

1. Deployment to a Git repository is greatly simplified. 
   Update your repository with results from the CI build.

1. Support for R packages and other kinds of project (bookdown, blogdown, etc.), with predefined templates. 
   Set up your project to deploy rendered versions of your book or blog with a single push to Git.

1. Workflow specification in a single `.R` file, regardless of CI system used.  
   Forget about `.yml` files or web browser configurations.
   
Other minor advantages include the use of `rcmdcheck::rcmdcheck()` for package checking a and robust caching approach of dependencies.

A more comprehensive summary of all advantages can be found in the [advantages](advantages.html) vignette.

# Initialization

The easiest way to use *tic* for CI services is to call `usethis::use_ci()`.
This will create templates for both "Travis CI" and "Appveyor" and initialize all the required authentication steps for deployment.
During the process, browser pages will open up, ensuring that all permissions are set correctly and all apps are authorized.
Also a Personal Access Token (PAT) will be created on GitHub and stored  as an encrypted variable on Travis.
This avoids hitting rate limitations when accessing the GitHub API, see [this blog post](https://itsalocke.com/blog/using-travis-make-sure-you-use-a-github-pat/) for more details.
The PAT creation cannot be fully automated as some input from the user on GitHub is required when creating the PAT.

First, authentication with Travis (a browser window opens) and GitHub is initiated (another browser window opens).
Also, the repo is actived on Travis.

```
usethis::use_ci()
✔ Setting active project to '/home/pjs/git/tic.package'
✔ GitHub is already initialized
Authenticating with GitHub
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
Authenticating with Travis
Authenticating with GitHub
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
Finished activating repo pat-s/tic.package on Travis CI.
```

Next, you will be asked if the current `.travis.yml` should be replaced with the template from *tic* (if one exists).
Otherwise, the template is added without asking.

```
Overwrite pre-existing file '.travis.yml'?
1: Absolutely not
2: Nope
3: Yeah
```

The same is done for Appveyor.

```
Selection: 3
✔ Writing '.travis.yml'
Overwrite pre-existing file 'appveyor.yml'?
1: Yeah
2: Nope
3: No way

Selection: 1
✔ Writing 'appveyor.yml'
```

And also for the `tic.R` file.
Additionally, it is added to `.Rbuildignore` to not interfer with the package building.

```
Overwrite tic.R? 

1: Yes
2: No

Selection: 1
Added tic.R from template.
✔ Adding '^tic\\.R$' to '.Rbuildignore'
```

Next, the private deploy key for Travis is added to the GitHub repo.
This is needed to give Travis permissions to deploy into the GitHub repo.

```
Authenticating with GitHub
Waiting for authentication in browser...
Press Esc/Ctrl + C to abort
Authentication complete.
Finished adding deploy keys on GitHub for repo pat-s/tic.package.
Successfully added public deploy key 'travis+tic for pat-s/tic.package' to GitHub for 
pat-s/tic.package. You should receive a confirmation e-mail from GitHub. Delete the 
key in the repository's settings to revoke access for that key or
when you no longer need it.
```

Besides the private deploy key, also the public deploy key needs to be added as an environment variable to Travis.
As it is a SSH key, its named `id_rsa` by default.

```
Finished adding private environment variable id_rsa to pat-s/tic.package on Travis CI.
Successfully added private deploy key to pat-s/tic.package as secure environment variable
id_rsa to Travis CI.
```

Last, a PAT is created to avoid the download limit of GitHub packages on Travis.
You now see two environment variables in your setting on Travis CI: `GITHUB_PAT` and `id_rsa`.

```
Create a personal access token, make sure that you are signed in as the correct user. 
The suggested description 'travis+tic for pat-s/tic.package' has been copied to the 
clipboard. If you use this token only to avoid GitHub's rate limit, you can leave 
all scopes unchecked. Then, copy the new token to the clipboard, it will be 
detected and applied automatically. Please visit
  https://github.com/settings/tokens/new
A browser window will be opened.
Waiting for PAT to appear on the clipboard.
Detected PAT, clearing clipboard.
Finished adding private environment variable GITHUB_PAT to pat-s/tic.package on Travis CI.
```

All this functionality is integrated in the [*usethis*](http://usethis.r-lib.org/) package because it contains various other useful `use_*` functions that simplify R (package) development.
However, the heavy lifting in the background is actually done by the R packages *travis* and *tic* . 
See [here](tic-usethis-travis.html) for more detailed information on how *tic* , *travis* and *usethis* work together.

# Explanation of the basic template

After having called `usethis::use_ci()` you will find a `.travis.yml`, `appveyor.yml` and a `tic.R` file in your repo.
Usually you do not need to touch `appveyor.yml` and `.travis.yml` anymore.
All build customizations are done in `tic.R` and apply to both services.
If you want more information about the whole build lifecycle, check the [Build lifecycle](build-lifecycle.html) vignette.

The basic `tic.R` template looks as follows:

```{r eval = FALSE}
do_package_checks(warnings_are_errors = (getRversion() >= "3.2"))

if (Sys.getenv("BUILD_PKGDOWN") != "") {
  get_stage("before_deploy") %>%
    add_step(step_setup_ssh())

  get_stage("deploy") %>%
    add_step(step_build_pkgdown()) %>% 
    add_step(step_push_deploy())
}
```

Let's break down what happens here:

```{r eval = FALSE}
do_package_checks(warnings_are_errors = (getRversion() >= "3.2"))
```

Is a macro which adds essential steps to various stages of a CI run. 
For example, it adds `step_rcmdcheck()` to the "script" stage. 
Check `?do_package_checks()` to see a list of all added steps by this convenience function.

```{r eval = FALSE}
if (Sys.getenv("BUILD_PKGDOWN") != "") { }
```

This line conditions the contents of the block inside the braces `{ }` on the existence of the environment variable `"BUILD_PKGDOWN"`.
The env variable is set in the [`.travis.yml`](https://github.com/ropenscilabs/tic/blob/d3c9ffab7e42ef4bf0a3113337c42dbaf592146c/.travis.yml#L45-L47) template, but only for the job that runs on the current R-release version:

```{yml eval = FALSE}
- r: 3.3
- r: 3.4
- r: release
  env:
    - BUILD_PKGDOWN=true
```

When the `if`-condition evaluates to `TRUE`, the following gets executed:  
First, `step_setup_ssh()` sets up the SSH connection needed for deployment.

```{r eval = FALSE}
get_stage("before_deploy") %>%
  add_step(step_setup_ssh())
```

Then, in the "deploy" stage of the CI run, function `step_build_pkgdown()` will build a [_pkgdown_](http://pkgdown.r-lib.org/) site if the project is an R package, and the `step_push_deploy()` step will push the updated files (if any) to Git.

```{r eval = FALSE}
get_stage("deploy") %>%
  add_step(step_build_pkgdown()) %>% 
  add_step(step_push_deploy())
```

# Examples projects

*tic* can be used for various R projects:

- [tic.package](https://github.com/ropenscilabs/tic.package): R packages with [_pkgdown_](https://pkgdown.r-lib.org/) documentation
- [tic.packagedocs](https://github.com/ropenscilabs/tic.packagedocs): R packages with [_packagedocs_](http://hafen.github.io/packagedocs/) documentation
- [tic.drat](https://github.com/ropenscilabs/tic.drat): CRAN-like package repositories with [_drat_](http://dirk.eddelbuettel.com/code/drat.html)
- [tic.blogdown](https://github.com/ropenscilabs/tic.blogdown): Blogs with [_blogdown_](https://bookdown.org/yihui/blogdown/)
- [tic.bookdown](https://github.com/ropenscilabs/tic.bookdown): Books with [_bookdown_](https://bookdown.org/)
- [tic.website](https://github.com/ropenscilabs/tic.website): Websites with [_rmarkdown_](https://rmarkdown.rstudio.com/)
- [tic.figshare](https://github.com/ropenscilabs/tic.figshare): Deploying artifacts to [figshare](https://figshare.com/) (work in progress).
- [tic.covrpage](https://github.com/ropenscilabs/tic.covrpage): Unit test summary report.

As a show case, we explain a "blogdown" project in more detail.
[`blogdown`](https://bookdown.org/yihui/blogdown/) is an R package for publishing websites.
Under the hood, it uses the framework [HUGO](https://gohugo.io/) which gets installed by the respective `tic.R` [template](https://github.com/krlmlr/tic.blogdown/blob/975aedd43fec1dd55e8348eccfca2c7c5f663006/tic.R#L5) in the "install" section:

```{r eval = FALSE}
get_stage("install") %>%
  add_code_step(blogdown::install_hugo(), 
                prepare_call = remotes::install_github("rstudio/blogdown"))
```

Then the website is built and deployed.
The `blogdown::build_site()` function for websites is the equivalent to `pkgdown::build_site()` for R packages.

```{r eval = FALSE}
get_stage("deploy") %>%
    add_code_step(blogdown::build_site()) %>%
    add_step(step_push_deploy())
```

Steps and stages differ between projects (e.g. between a "blogdown" website and a "package").
*tic* is smart enough to detect your project automatically when calling `usethis::use_ci()` and will add the correct template.

**Note:** Currently, publishing to https://figshare.com/ doesn't work.
Also, publishing to https://zenodo.org/ is work in progress.

## *tic* based R projects from the community

The templates we provide with *tic* are minimal working examples.
In the following we list various R projects with non-standard/tailored *tic*-based workflows.
You may find inspiration for your own project there.
(Please contact us if you want to add yours to the list!)

### Packages

- [mlr](https://github.com/mlr-org/mlr)
- [sf](https://github.com/r-spatial/sf)
- [dbi](https://github.com/r-dbi/DBI)

### Bookdown

### Blogdown

# Advanced

The advanced usage of *tic* is described in more detail in the [Advanced Usage](advanced.html) article:

- [Options for _pkgdown_ deployment](advanced.html#pkgdown-deployment) 
- [Using Travis CI Meta-information](advanced.html#using-travis-ci-meta-information)
- [Troubleshooting: Running *tic* locally](advanced.html#troubleshooting-running-tic-locally)
- [Troubleshooting: Enter into the Travis build](advanced.html#troubleshooting-running-tic-locally)
- [Writing custom steps](custom-steps.html)

The build lifecycle when using *tic* is explained in the [Build lifecycle](build-lifecycle.html) article.
