---
title: "Troubleshooting"
author: "Patrick Schratz, Kirill Müller"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette lists common errors that one might encounter during all aspects of the CI world.

The grouping may change overtime depending on the number of issues we receive for certain sub-topics.

## Appveyor

```{r eval=FALSE}
Error in eval(expr, envir = list(`?` = function(...) stop()), enclos = envir) : 
  object 'name' not found
Calls: <Anonymous> ... .transformer -> glue -> glue_data -> <Anonymous> -> .transformer
Execution halted
Command exited with code 1
```

Do you have a private SSH key stored as a secure environment variable in the in `appveyor.yml`?
If yes, it could be that you encrypted while being logged in to a different account than the one the repo is running under in reality.
This means if you have access to multiple organizations and your private account on Appveyor, you need to encrypt the environment variable using the account in which repo lives.
